# Moving Pattern Analysis of Data Collected from Personal Activity Monitoring Device

This is an educational project, which is the part of [Data Science Specialization](https://www.coursera.org/specialization/jhudatascience/1?utm_medium=listingPage) provided by Johns Hopkins University on Coursera.

## Introduction

It is now possible to collect a large amount of data about personal
movement using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or
[Jawbone Up](https://jawbone.com/up). These type of devices are part of
the "quantified self" movement - a group of enthusiasts who take
measurements about themselves regularly to improve their health, to
find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for
processing and interpreting the data.

## Data

The data for this assignment can be downloaded from the [Coursera](https://www.coursera.org/) web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

It consists of data from personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this dataset.

Data was loaded and transformed to appropriate format with correctly represented date-time.

```{r loaddata, cache=TRUE}
if (!file.exists("activity.csv")) {
    unzip("activity.zip", overwrite=FALSE)
}
data <- read.table("activity.csv", header=TRUE, sep=",", 
                   colClasses=c("numeric", "character", "numeric"), 
                   na.strings="NA", nrows=17569, allowEscapes=TRUE, quote='"')
data$date <- as.Date(data$date, format="%Y-%m-%d")
str(data)
```



## What are the statistical parameters of the number of steps taken per day?

In order to obtain mean and median number of steps per day was summed up data 
for all 5 minute intervals for each date. Then mean and median were calculated and 
was built a histogram of the total number of steps taken per day in October and 
November 2012.

```{r daysummary1}
day_summary <- tapply(data$steps, data$date, sum, na.rm=TRUE)
mean(day_summary)
median(day_summary)
```

```{r globaverage, fig.height=6, fig.width=6}
hist(day_summary, breaks=15, xlim=c(0, 25000), ylim=c(0, 20), cex.lab=1.3,  
     xlab="Number of steps per day", col="steelblue",  
     main=paste("Histogram of the total number of steps",  
                "per day\n from anonymous individual. \n",  
                "Collected in October-November, 2012"))
```



## What is the average daily activity pattern?

The average number of steps taken for each 5 minute time interval throughout a 
day was calculated.

```{r dayaverage, fig.height=6, fig.width=6}
time_summary <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
plot(as.numeric(names(time_summary)), time_summary, type="l", cex.lab=1.3,  
     ylab="Average number of steps", xlab="Day timeline, minutes", 
     col="steelblue",  
     main=paste("Average number of steps throughout the day",  
                "from \n anonymous individual. Collected in",  
                " \nOctober-November, 2012"))
```

Then we found the 5 minute time interval where the number of steps is maximal 
on average throughout the day.

```{r maxsteps}
max_steps <- as.numeric(names(which.max(time_summary)))
max_steps_hours <- max_steps %/% 60
max_steps_minutes <- max_steps %% 60
```

The maximal average number of steps was for `r max_steps` minutes time interval, 
which corresponds time between `r max_steps_hours`:`r max_steps_minutes - 5` 
and `r max_steps_hours`:`r max_steps_minutes`.

## Imputing missing values

The total number of missing values in the dataset (i.e. the total number of rows 
with NAs) was calculated.

```{r NAnum}
sum(!complete.cases(data))
```

There are 2304 rows with NA in dataset. 

In order to impute NA values we created a new dataset that was equal to the 
original dataset but with the missing data filled in with mean for 5-minute 
interval that corresponds to given missing value.

```{r removeNA}
data_filled <- data
for (x in 1:length(data_filled$steps)) {
    if (is.na(data_filled$steps[x])) {
        data_filled$steps[x] = time_summary[as.character(data_filled$interval[x])][[1]]
    }
}
head(data)
head(data_filled)
```

After NA imputaion we calculated mean, median again, and built a histogram of 
the total number of steps per day, but this time for dataset with filled NAs. 

```{r daysummary2}
day_summary <- tapply(data_filled$steps, data_filled$date, sum, na.rm=TRUE)
mean(day_summary)
median(day_summary)
```

```{r globalsteps, fig.height=6, fig.width=6}
hist(day_summary, breaks=15, cex.lab=1.3, ylim=c(0, 25), xlim=c(0, 25000),  
     xlab="Number of steps per day", col="slategrey",  
     main=paste("Histogram of the total ",  
                "number of steps per day from ",  
                "anonymous\n individual, collected in October-November, 2012 "))
```

As we can see, this graph differ significantly from the histogram for global 
average number of steps without removed NAs. Now there are no peak at 0 in histogram and
we have significantly higher main peak. Also removing NA makes mean and median equal, 
which means that data no longer skewed.

## Are there differences in activity patterns between weekdays and weekends?

At the next stage we exemined if the pattern of number of steps is differ for 
weekdays and weekends. At first we created a new variable "weekday", which is a 
factor with two levels “weekday” and “weekend”.

```{r addweekday}
weekday <- factor(as.POSIXlt(data_filled$date)$wday == 0 | as.POSIXlt(data_filled$date)$wday == 6,  
                  labels=c("weekday", "weekend"))
data_filled$weekday <- weekday
str(data_filled)
```

After that we made a panel plot containing a time series plot of the 5-minute 
interval and the average number of steps taken, averaged across all weekday 
days or weekend days.

```{r timeseries}
head(data_filled)
agg <- aggregate(data_filled$steps, 
                 by=list(data_filled$interval, data_filled$weekday), FUN=mean)
colnames(agg) <- c("interval", "weekday", "steps")
head(agg)
library(lattice)
xyplot(steps ~ interval | weekday, data=agg, layout=c(1, 2), type="l",  
       xlab=list(label="Time, minutes", cex=1.5),  
       ylab=list(label="Average number of steps", cex=1.5), 
       main=paste("Distribution of the average number of steps throughout ",  
                  "the day\n for anonymous individual. ",  
                  "Collected in October-November, 2012"))
```

As we can see, there are different moving patterns for weekend and weekdays. 
On weekdays average number of steps is higher in the morning and lower in the 
rest of the day comparing to weekends.


The data was kindly provided by Data Science Specialization courses, Coursera.